---
title: "BIGf90 Tutorial"
subtitle: "Variance components, breeding values, and cross-validation with the BLUPf90 programs"
author: "Breeding Insight - Josue Chinchilla-Vargas"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    css: BIcolors_template.css
    includes:
      in_header: header.html
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{BIGf90 Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,          # steps call the external BLUPf90 programs, so code is shown but not run
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

# Introduction

BIGf90 is an R package developed by [Breeding Insight](https://breedinginsight.org/)
that wraps the [BLUPf90 family of programs](http://nce.ads.uga.edu/wiki/doku.php)
for genetic evaluation. This tutorial walks through a full analysis:

* `write_par()` &mdash; build the RENUMF90 parameter file from R.
* `run_renum()` &mdash; renumber the model with RENUMF90.
* `run_gibbs()` / `run_postgibbs()` &mdash; estimate variance components.
* `run_blup()` / `clean_ebvs()` &mdash; compute (genomic) breeding values.
* `bf90_cv()` &mdash; K-fold cross-validation (predictive ability, bias, accuracy).

The same pipeline runs with or without genotypes (GBLUP/ssGBLUP vs pedigree BLUP).
Because these functions call the external BLUPf90 executables, the code below is
shown but not executed when the vignette is built &mdash; run it locally with the
programs installed.

## Install

```{r}
install.packages("remotes")
remotes::install_github("Breeding-Insight/BIGf90", dependencies = TRUE)
library(BIGf90)
```

# What you need

* The BLUPf90 executables (`renumf90`, `blupf90+`, `predictf90`, `gibbsf90+`,
  `postgibbsf90`) in one folder &mdash; download from the
  [BLUPF90 wiki](http://nce.ads.uga.edu/wiki/doku.php).
* A phenotype file, a pedigree file, and (optionally) a genotype/SNP file.

# Setup

```{r}
library(BIGf90)

path_2_execs <- "/path/to/blupf90/executables/"   # folder with the BLUPf90 programs
base_dir     <- "/path/to/your/analysis"          # your data + parameter files
setwd(base_dir)

animal_effect <- 3       # animal effect's NUMBER in the model (for clean_ebvs / bf90_cv)
h2            <- 0.3     # narrow-sense heritability estimate (for CV accuracy)
missing_code  <- -999    # missing-phenotype code
```

Two "animal" indices, easy to mix up: in `write_par()` below, `random` is the animal
effect's **column in the data file**; in `clean_ebvs()`/`bf90_cv()`, `random_effect_col`
is its **number in the model**. Usually different values.

# Prepare genotypes (optional)

For a genomic model you need a BLUPf90 genotype file: each line is the animal ID
followed by a contiguous string of one 0/1/2 dosage per SNP. If you are starting
from PLINK, `write_geno()` builds it directly from a `.ped`. By default it reads
the `--recode12` layout (two allele codes per locus) and collapses each pair to a
dosage (1 1 &rarr; 0, 1 2 &rarr; 1, 2 2 &rarr; 2, and any `0` &rarr; missing),
replacing the usual `cut`/`tr`/`sed` reformatting. Skip this step if you already
have a `.geno` file or are running a pedigree-only model.

```{r}
write_geno(
  ped  = "ped012.ped",     # PLINK --recode12 output (no header, 6 lead columns)
  file = "bf90_geno.txt",  # BLUPf90 genotype file to create
  map  = "ped012.map",     # optional: checks the locus count matches
  id_col = 2               # column with the animal ID (1 = FID, 2 = IID)
)
```

# Build the parameter file

`renumf90` needs a parameter file describing the model. Build it from R (or write it
by hand). Omit `snp_file` and the genomic `OPTION`s for a pedigree-only model.

```{r}
write_par(
  file     = "model.par",
  datafile = "phenotypes.txt",
  traits   = 4,                          # column of the trait in the data file
  residual_variance = 0.15,
  effects  = list(
    mean   = list(col = 2, type = "cross", class = "numer"),
    fixed1 = list(col = 3, type = "cross", class = "numer"),
    animal = list(col = 1, type = "cross", class = "alpha")   # the animal effect
  ),
  random        = 1,                     # animal effect's column in the data file
  random_type   = "animal",
  pedigree_file = "pedigree.ped",
  snp_file      = "genotypes.geno",      # omit for a pedigree-only model
  missing_value = missing_code,
  options = c("callrate 0.9", "minfreq 0.05", "monomorphic 1",
              "saveGInverse", "saveCleanSNPs")
)
```

# Renumber the model

```{r}
prep_renum <- function(par_file, out_dir) {
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  run_renum(path_2_execs, raw_par_file = file.path(base_dir, par_file),
            output_files_dir = out_dir, verbose = FALSE)
  inb <- file.path(base_dir, "renf90.inb")
  if (file.exists(inb)) file.rename(inb, file.path(out_dir, "renf90.inb"))
}

out <- file.path(base_dir, "renumbered")
prep_renum("model.par", out)
```

# Variance components

```{r}
run_gibbs(path_2_execs, input_files_dir = out, output_files_dir = out,
          gibbs_iter = 100000, gibbs_burn = 20000, gibbs_keep = 10, verbose = FALSE)
run_postgibbs(path_2_execs, input_files_dir = out,
              postgibbs_burn = 1, postgibbs_keep = 1, verbose = FALSE)
```

# Breeding values

```{r}
setwd(out)                 # run_blup / clean_ebvs operate on the current folder
run_blup(path_2_execs)
clean_ebvs(random_effect_col = animal_effect,
           solutions_output_name = "ebvs_clean.txt")
setwd(base_dir)
```

`ebvs_clean.txt` holds the (G)EBVs keyed back to the original animal IDs.

# Cross-validation

```{r}
set.seed(101919)   # repeatable folds
bf90_cv(path_2_execs,
        missing_value_code = missing_code,
        random_effect_col  = animal_effect,
        h2                 = h2,
        num_runs           = 10,     # independent CV replicates
        num_folds          = 5,      # folds per replicate
        input_files_dir    = out,
        output_files_dir   = out,
        output_table_name  = "cv_results.txt",
        verbose            = FALSE)
```

`cv_results.txt` reports, using the phenotype corrected for fixed effects (y\*):
**predictive ability** = cor(y\*, êbv), **bias** = the slope of the regression of
y\* on êbv (1 = unbiased), and **accuracy** = predictive ability / √h².

# Genomic vs pedigree-only

Run the pipeline twice into separate folders to compare:

* **Genomic:** keep `snp_file` + `saveGInverse`/`saveCleanSNPs` in the `.par`.
* **Pedigree-only:** drop `snp_file` and the genomic `OPTION`s.

Then compare the `cv_results.txt` / `ebvs_clean.txt` between the two.

# Session information

```{r, eval=TRUE}
sessionInfo()
```

# How to cite

If you use BIGf90 in research, run `citation("BIGf90")` for the current reference,
and cite the BLUPf90 programs (Misztal et al. 2014).

```{r, eval=TRUE}
citation("BIGf90")
```

# Acknowledgments

Developed as part of the [Breeding Insight](https://www.breedinginsight.org)
initiative to provide open-source, data-driven tools for modern breeding programs.
